---
title: 'EWGT-2021: Potential of vision-enhanced floating car data for urban traffic
  estimation'
author: "Dmitry Pavlyuk"
date: "December 5, 2020"
output: pdf_document
editor_options:
  chunk_output_type: console
---

This markdown reproduces the research "Potential of vision-enhanced floating car data for urban traffic estimation, submitted to the 24rd EURO Working Group on Transportation Meeting, EWGT 2021.

Load necessary libraries
----------------
```{r libs}
require(needs)
needs(igraph)
needs(matrixcalc)
needs(fsMTS)
needs(plot.matrix)
needs(expm)
needs(dplyr)
needs(ggplot2)
needs(tidyr)
needs(scales)
source("functions.R")
```
Load necessary libraries
----------------
```{r paths}
data.folder <- file.path(getwd(),"data")
output.folder <- file.path(getwd(),"output")
```

Prepare PeMS data
```{r sampling}
pems.rds <- file.path(data.folder, "PeMS", "prepared.rds")
if (!file.exists(pems.rds)) {
  meta <-
    read.csv(
      file.path(data.folder, "PeMS", "d07_text_meta_2020_11_16.txt"),
      header = T,
      sep = "\t"
    )
  stations <-
    c(
      717046,
      717045,
      717263,
      717264,
      716943,
      716942,
      716331,
      717445,
      717047,
      716028,
      716946,
      718085,
      718173,
      716939
    )
  observable <-
    list(
      '717046' = '717045',
      '717045' = '717046',
      '717263' = '717264',
      '717264' = '717263',
      '716943' = '716942',
      '716942' = '716943',
      '716331' = '717445',
      '717445' = '716331',
      '717047' = '716028',
      '716028' = '717047',
      '716946' = '718085',
      '718085' = '716946',
      '718173' = '716939',
      '716939' = '718173'
    )
  data.raw <-
    read.csv(file.path(
      data.folder,
      "PeMS",
      "d07_text_station_5min_2020_11_29.txt"
    ),
    header = F)
  data.tb <- as_tibble(data.raw)
  data.tb <-
    data.tb %>% mutate(
      datetime = as.POSIXct(V1, format = "%m/%d/%Y %H:%M:%S"),
      station = V2,
      volume = V10,
      occupancy = V11,
      speed = V12
    ) %>%
    select(datetime, station, volume, occupancy, speed)
  data.tbf <-
    data.tb %>% filter(station %in% stations) %>% 
    mutate(speed = ifelse(is.na(speed), 65, speed)) %>%
    
    mutate(volume = ifelse(is.na(volume), 0, volume))
  
  
  vols <-
    data.tbf %>% select(datetime, station, volume) %>%
    pivot_wider(names_from = station, values_from = volume)
  rels <- cor(vols %>% select(-datetime))
  md <- rowSums(abs(rels))
  Lw <- -abs(rels) + diag(ncol(rels)) + diag(md)
  data.prepared <-
    list(
      traffic.data = data.tbf %>% mutate(station = as.factor(station)) %>%
        select(datetime, station, speed, volume),
      Lw = Lw,
      observable = observable
    )
  saveRDS(data.prepared, file = pems.rds)
} else{
  warning("Prepared data exists: loading")
  data.prepared <- readRDS(file = pems.rds)
}
```

# Descriptive analysis
```{r}
data.speed <-
  data.prepared$traffic.data %>% select(datetime, station, speed) %>%
  pivot_wider(names_from = station, values_from = speed) %>% select(-datetime)
data.volume <-
  
  data.prepared$traffic.data %>% select(datetime, station, volume) %>%
  pivot_wider(names_from = station, values_from = volume) %>% select(-datetime)
N <- ncol(data.speed)
T <- nrow(data.speed)
```
Data dimensions: `r dim(data.speed)`

Speed plots
```{r}  
data.prepared$traffic.data %>% ggplot(aes(
  x = datetime,
  y = speed,
  col = station,
  group = station
)) + geom_line(size = 1)

```

Volume plots
```{r}  
data.prepared$traffic.data %>% ggplot(aes(
  x = datetime,
  y = volume,
  col = station,
  group = station
)) + geom_line(size = 1)
```

Speed variance
```{r}
data.prepared$traffic.data %>% group_by(station) %>% summarise(sd(speed))
```

# Traffic estimation

```{r}
sp <- 5e-4
omega <- randomOmega(N, T, sp, t(data.volume))
sum(omega > 0) / (N * T)
res <-
  TGMCS(t(data.speed),
        data.prepared$Lw,
        omega,
        returnQhat = T,
        lambda3 = 10)
hat.tbf <- as_tibble(t(res$Qhat))
hat.tbf$datetime <-
  unique(data.prepared$traffic.data %>% pull(datetime))
hat.tbf %>% pivot_longer(-one_of("datetime"),
                         names_to = "station",
                         values_to = "speed") %>%
  ggplot(aes(
    x = datetime,
    y = speed,
    col = station,
    group = station
  )) + geom_line(size = 1)
res$observedMAE
res$unobservedMAE
res$observedMAPE
res$unobservedMAPE
```

# Experiments
```{r}
results.rds <- file.path(output.folder, "results.rds")
sp.list <- c(4e-5, 1e-4, 2e-4, 3e-4, 4e-4, 5e-4, 7e-4, 9e-4)
sp.list <- c(2e-4, 3e-4, 4e-4, 5e-4, 7e-4, 9e-4, 13e-4)
for (sp in sp.list) {
  omega <- randomOmega(N, T, sp, t(data.volume))
  print(paste(sp, "=", round(sum(omega > 0) / (N * T), 2)))
}
if (!file.exists(results.rds)) {
  est <- list()
  r <- 15
  mu = 0.0001
  lambda1 = 0.01
  lambda2 = 0.05
  lambda3 = 0.1
  tol = 1e-6
  maxIter = 1e6
  for (rep in 1:50) {
    for (sp in sp.list) {
      print(paste(rep, sp))
      omega <- randomOmega(N, T, sp, t(data.volume))
      omegax2 <- omega + randomOmega(N, T, sp, t(data.volume))
      omegax2[omegax2 > 1] <- 1
      omegaExt <- enhanceOmega(omega, data.prepared$observable)
      est[[length(est) + 1]] <-
        c(
          TGMCS(
            t(data.speed),
            data.prepared$Lw,
            omega,
            accMask = omega,
            maxIter = maxIter,
            tol = tol,
            r = r,
            mu = mu,
            lambda1 = lambda1,
            lambda2 = lambda2,
            lambda3 = lambda3
          ),
          sparsity = sp,
          obslinks = sum(omega > 0),
          coverage = sum(omega > 0) / (N * T),
          name = "omega"
        )
      est[[length(est) + 1]] <-
        c(
          TGMCS(
            t(data.speed),
            data.prepared$Lw,
            omegax2,
            accMask = omegax2,
            maxIter = maxIter,
            tol = tol,
            r = r,
            mu = mu,
            lambda1 = lambda1,
            lambda2 = lambda2,
            lambda3 = lambda3
          ),
          sparsity = sp,
          obslinks = sum(omegax2 > 0),
          coverage = sum(omegax2 > 0) / (N * T),
          name = "omegax2"
        )
      est[[length(est) + 1]] <-
        c(
          TGMCS(
            t(data.speed),
            data.prepared$Lw,
            omegaExt,
            accMask = omegaExt,
            maxIter = maxIter,
            tol = tol,
            r = r,
            mu = mu,
            lambda1 = lambda1,
            lambda2 = lambda2,
            lambda3 = lambda3
          ),
          sparsity = sp,
          obslinks = sum(omegaExt > 0),
          coverage = sum(omegaExt > 0) / (N * T),
          name = "omegaExt"
        )
      print(tail(bind_rows(est)))
    }
    saveRDS(est, file = results.rds)
  }
} else{
  est <- readRDS(results.rds)
}
```

Experimental results
```{r}
est.df <- bind_rows(est)
est.df

mCov.df <-
  est.df %>% filter(name == "omega") %>% group_by(sparsity) %>%
  summarise(meanCoverage = mean(coverage))
est.df <- est.df %>% left_join(mCov.df, by = c("sparsity"))

f <- 1.96
legend_names <-
  c('omega' = 'FCD',
    'omegax2' = 'FCD x 2',
    'omegaExt' = 'vision-xFCD')
est.df %>% filter(converged == TRUE, sparsity > 1e-4, sparsity <= 13e-4) %>%
  mutate(name = legend_names[name]) %>%
  group_by(name, sparsity, meanCoverage) %>%
  summarise(
    meanMAE = mean(unobservedMAE),
    sdMAE = sd(unobservedMAE),
    n = n(),
    minMAE = min(unobservedMAE),
    maxMAE = max(unobservedMAE),
    lb = max(meanMAE - f * sdMAE / sqrt(n), minMAE),
    ub = min(meanMAE + f * sdMAE / sqrt(n), maxMAE)
  ) %>%
  ggplot(aes(
    x = meanCoverage,
    y = meanMAE,
    col = name,
    group = name,
    linetype = name
  )) + geom_line(size = 1.2) +
  geom_ribbon(aes(
    ymin = lb,
    ymax = ub,
    col = name,
    group = name,
    linetype = name
  ),
  alpha = 0.1) +
  scale_x_continuous(
    labels = function(x)
      paste0(x * 100, "%")
  ) + labs(x = "% of observed road segments", y = "Mean MAE, mph")

est.df %>% filter(converged == TRUE, sparsity > 1e-4, sparsity <= 13e-4) %>%
  mutate(name = legend_names[name]) %>%
  group_by(name, sparsity, meanCoverage) %>%
  summarise(
    meanMAPE = mean(unobservedMAPE),
    sdMAPE = sd(unobservedMAPE),
    n = n(),
    minMAPE = min(unobservedMAPE),
    maxMAPE = max(unobservedMAPE),
    lb = max(meanMAPE - f * sdMAPE / sqrt(n), minMAPE),
    ub = min(meanMAPE + f * sdMAPE / sqrt(n), maxMAPE)
  ) %>%
  ggplot(aes(
    x = meanCoverage,
    y = meanMAPE,
    col = name,
    group = name,
    linetype = name
  )) + geom_line(size = 1.2) +
  geom_ribbon(aes(
    ymin = lb,
    ymax = ub,
    col = name,
    group = name,
    linetype = name
  ),
  alpha = 0.05) +
  scale_y_continuous(
    labels = function(x)
      paste0(x * 100, "%")
  ) +
  scale_x_continuous(
    labels = function(x)
      paste0(x * 100, "%")
  ) + labs(x = "% of observed road segments", y = "Mean MAPE, %")

est.df %>% filter(converged == TRUE, sparsity > 1e-4, sparsity <= 13e-4) %>%
  mutate(name = legend_names[name]) %>%
  group_by(name, sparsity, meanCoverage) %>%
  summarise(meanMAPE = mean(unobservedMAPE) * 100,
            meanSp = mean(obslinks) / (N * T)) %>%
  mutate(coverage = paste0(round(meanCoverage * 100), "%")) %>%
  pivot_wider(
    id_cols = c(name),
    names_from = "coverage",
    values_from = "meanMAPE"
  )
```